mathml: Translate R expressions to MathML and LaTeX

### 2023-01-21 rj/mathml.Rmd
intro
pdf

---
title: "mathml: Translate R expressions to MathML and LaTeX"
date: "2023-01-21"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the RMarkdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package allows to use the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks allow to extend the package by
  mapping specific variables or functions to new MathML and LaTeX entities. A
  few working examples illustrate the use of the package in a scientific report.
draft: true
author:  
  - name: Matthias Gondan
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: bibliography.bib
preamble:
- \usepackage{cancel}
- \usepackage{amsmath}
Vignette: >
  %\VignetteIndexEntry{mathml: Translate R expressions to MathML and LaTeX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

@@

The vignette is organized as follows. Section\ 2 presents the technical
background of the package, including the two main classes of Prolog rules for
translating R\ objects to mathematical expressions. Section\ 3 illustrates the
main features of the \pkg{mathml} package, potential issues and workarounds
using examples from the day-to-day perspective of a user. Section\ 4 present a
case study for the use of the package in a scientific report. Section\ 5
concludes with a discussion and ideas for further development.

@@

The translation of R\ expressions to mathematical output is achieved through a
Prolog interpreter. Prolog is a classical logic programming language with many
applications in expert systems, computer linguistics and symbolic artificial
intelligence. The main strength of Prolog is its concise representation of facts
and rules for the representation of knowledge and grammar, as well as its
efficient built-in search engine for closed world domains. As it is well-known,
R\ is a statistical programming language for data analysis and statistical
modeling which is widely used in academia and industry. Besides the core
library, a lot of packages have been developed for all kinds of statistical
problems, including statistics-based artificial intelligence tools such as
neural networks for machine learning and deep learning. Whereas Prolog is weak
in statistical computation, but strong in symbolic manipulation, the converse
may be said for the R\ language. The \pkg{rolog} package [@rolog] bridges this
gap by providing an interface to a SWI-Prolog distribution in R. The
communication between the two systems is mainly in the form of queries from
R\ to Prolog, but two predicates allow Prolog to ring back and evaluate terms in
R.

@@

As is seen from the output above, the variable `term` is not assigned the result
of the calculation, but an R\ "call"
[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. This call
can eventually be evaluated with `eval()`,

@@

We can include the output in a RMarkdown document by specifying `results='asis'`
in the R\ code chunk, as is done in the next example. The
R\ function `mathout()` is a wrapper that invokes `mathml()` for HTML output
and `mathjax()` for LaTeX output.

@@

The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
with `N` representing the name of the function and `X` its arguments and
parameters, enclosed in parentheses. A corresponding rule `jax/3` does the same
for MathJax/LaTeX. A list of flags can be used for context-sensitive
translation (see, e.g., the errors below).

@@

An expression such as `1 + -2` may be considered unsatisfactory from an
aesthetical perspective. It is correct R\ syntax, though, and is reproduced
accordingly, without the parentheses. Parentheses around negated numbers or
symbols can be added as shown for `-Inf`.

Note that an R function `hat()` does not exist in base R, it is provided by the
package for convenience and points to the identity function.

@@

For multiplications involving only numbers and symbols, the multiplication sign
is omitted. This heuristic does not always produce the desired result;
therefore, \pkg{mathml} defines alternative multiplication
functions `dot()`, `nodot()`, and `times()` that calculate the product and
produce the respective multiplication signs. Similarly, `frac()` and `dfrac()`
can be used for small and large fractions.

@@

```{r, results='asis'}
term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
mathout(term)
term <- quote('%+-%'(mean(X), 1.96 * s / sqrt(N)))
term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
mathout(term)
```

The example is a reminder that it is not possible to define the precedence of
custom operators in R, and that expressions with such operators are evaluated
strictly from left to right. As in the previous example, the solution is to
either use the functional form of the operator or a curly brace to enforce the
correct operator precedence.

@@

There is support for most functions from \pkg{base}, with adequate use and omission
of parentheses.

```{r, results='asis'}
term <- quote(sin(x) + sin(x)^2L + cos(pi/2L) + tan(2L*pi) * expm1(x))
mathout(term)
term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + ceiling(x))
mathout(term)
```

@@

```{r base-stats, echo=FALSE}
op1 <- list(
  "sin(x)"=quote(sin(x)),
  "cosh(x)"=quote(cosh(x)),
  "tanpi(alpha)"=quote(tanpi(alpha)),
  "asinh(alpha)"=quote(asinh(alpha)),
  "log(p)"=quote(log(p)),
  "log1p(x)"=quote(log1p(x)),
  "logb(x,\\ e)"=quote(logb(x, e)),
  "exp(x)"=quote(exp(x)),
  "expm1(x)"=quote(expm1(x)),
  "choose(n,\\ k)"=quote(choose(n, k)),
  "lchoose(n,\\ k)"=quote(lchoose(n, k)),
  "factorial(n)"=quote(factorial(n)), 
  "lfactorial(n)"=quote(lfactorial(n)),
  "sqrt(x)"=quote(sqrt(x)),
  "mean(X)"=quote(mean(X)),
  "abs(x)"=quote(abs(x)))
m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))

@@

For self-written functions, matters are a bit more complicated. The name of a
self-written function is not transparent in R, since only the function body is
represented, as illustrated by the redefinition of `sign` below. The function is
often a nested R "call" of the form `'{'(L)`, with `L` being a list of
commands. \pkg{mathml} provides very limited support for typical control
structures.

@@

term <- quote(mysign <- sgn(x))
mathout(term)

hook(mysign, sg(x))
term <- call("<-", quote(mysign), sgn)
mathout(term)
```

The hook (see below) is used to display \(\mathrm{sg}(x) = \{\ldots\) instead
of just \(s = \{\).

@@

For mathematical typesetting in the form
of \(\int f(x)\, dx\), \pkg{mathml} needs to find out the name of the
integration variable. For that purpose, the underlying Prolog bridge provides a
predicate `r_eval/2` that calls R\ from Prolog. In the example above, this
predicate evaluates `formalArgs(args(sin))`, which returns the names of the
arguments of `sin`, namely, `x`.

Note that in the example above, the quoted term is an abbreviation
for `call("integrate", quote(sin), ...)`, with `sin` being an R\ symbol, not a
function. R's `integrate()` can handle both. However, as already mentioned in
the previous subsection, \pkg{mathml} would be unable to find out the function
name, it needs the symbol.

@@

One of R's great features is the possibility to refer to function arguments by
their names, not only by their position in the list of arguments. At the other
end, most Prolog-powered handlers for the R\ calls are rather rigid, for
example, `integrate/3` accepts exactly three arguments in a particular order and
without names, that is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print
the desired result.

To "canonicalize" function calls with named arguments and arguments in unusual
order, \pkg{mathml} provides an auxilliary R\ function `canonical(f, drop)` that
reorders the argument list of calls to known R\ functions and, 
if `drop=TRUE` (which is the default), also removes the names of the arguments.

@@

Note that the seemingly more convient `term <- quote(A + B)` yields \(A + B\) in
the output---instead of the desired matrix representation.

@@

We consider the \(t\)-statistic for independent samples with equal variance. To
avoid clutter in the equation, the pooled variance \(s^2_{\mathrm{pool}}\) is
abbreviated, and a comment is given with the expression
for \(s^2_{\mathrm{pool}}\). For this purpose, \pkg{mathml} provides a
function `denote(abbr, expr, info)`, with `expr` actually being
evaluated, `abbr` being rendered, plus a comment of the
form "with `expr` denoting `info`". The hooks are described in the next section.

@@

The term is evaluated below. `print()` is needed because the return value of the
assignment in the term `t <- dfrac(...)` is invisible.

@@

```{r, results='asis'}
t <- quote(dfrac(omit_right(mean(D) - mu[0L]), s / sqrt(N)))
mathout(t, flags=list(error='highlight'))
mathout(t, flags=list(error='fix'))
```

@@

This case study describes a model by Schwarz [-@schwarz1994] from mathematical
Psychology using the features of package \pkg{mathml}. Schwarz presents a new
explanation of redundancy gains that occur when observers respond to stimuli of
different sources, and the same information is presented on two or more
channels. In Schwarz' model, decision-making builds on a process of noisy
accumulation of information over time [e.g., @ratcliff2016]. In redundant
stimuli, the model assumes a superposition of channel-specific diffusion
processes that eventually reach an absorbing barrier to elicit the response. For
a detailed description the reader may refer to the original article.

## The model

Schwarz' [-@schwarz1994] model refers to two stimuli A and B, presented either
alone or in combination (AB, redundant stimuli), with the redundant stimuli
being presented either simultaneously or with onset asynchrony \(\tau\). The
channel activation is described as a two-dimensional Wiener process with
drifts \(\mu_i\), variances \(\sigma^2_i\), and initial
conditions \(X_i(t = 0) = 0, i = \mathrm{A, B}\). The channels may be correlated
with \(\rho_{\mathrm{AB}}\), but we assume \(\rho_{\mathrm{AB}} = 0\) for
simplicity.

A response is elicited when the process reaches an absorbing barrier \(c > 0\)
for the first time. In single-target trials, the first passages at \(c\) are
expected at

@@

In redundant stimuli, the activation of the channel-specific diffusion processes
adds up, \(X_{\mathrm{AB}}(t) = X_{\mathrm A}(t) + X_{\mathrm B}(t)\), hence
the name, superposition. \(X_{\mathrm{AB}}(t)\) is again a Wiener process with
drift \(\mu_{\mathrm A} + \mu_{\mathrm B}\) and
variance \(\sigma^2_{\mathrm A} + \sigma^2_{\mathrm B}\). For the expected
first-passage time, we have

@@

```{r, results="asis"}
ED_async <- function(tau, c, mu_A, sigma_A, mu_B)
{ dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B)) *
    ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
      - (mu_A*tau + c) * exp(dfrac(2L*mu_A*c, sigma_A^2L))
        * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau))))
}

@@

For negative onset asynchrony (i.e., B before A), the parameters need to be
switched.

```{r, results='asis'}
ED <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
{
  if(tau == Inf) return(ED_single(c, mu_A)) ;
  if(tau == -Inf) return(ED_single(c, mu_B)) ;
  if(tau == 0L) return(ED_single(c, mu_A + mu_B)) ;
  if(tau > 0L) return(ED_async(tau, c, mu_A, sigma_A, mu_B)) ;
  if(tau < 0L) return(ED_async(abs(tau), c, mu_B, sigma_B, mu_A))
}

@@

## Methods

@@

```{r miller-data, echo=FALSE}
tau  <- list(-Inf, -167L, -133L, -100L, -67L, -33L, 0L, 33L, 67L, 100L, 133L, 167L, Inf)
m    <- c(231, 234, 230, 227, 228, 221, 217, 238, 263, 277, 298, 316, 348)
s    <- c( 56,  58,  40,  40,  32,  28,  28,  28,  26,  30,  32,  34,  92)
n    <- c(400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400)

@@

t <- cbind('$\\tau$'=ttau, '$m$'=tm, '$s$'=ts, '$n$'=tn)
knitr::kable(t, caption="Miller data", row.names=FALSE, escape=FALSE)
```

Assuming that the model is correct, the observable mean reaction times follow an
approximate Normal distribution around the model prediction \(E(T)\). We can,
therefore, use a standard goodness-of-fit measure by \(z\)-standardisation.

```{r, results='asis'}
z <- function(m, s, n, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
  dfrac(m - denote(E, ET(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M),
              "the expected mean response time"),
    s / sqrt(n))

@@

The overall goodness-of-fit is the sum of the squared \(z\)-statistics for each
onset asynchrony. Assuming again that the architecture of the model is correct,
but the parameters are adjusted to the data, it follows
a \(\chi^2(8~\mathrm{df})\)-distribution.

```{r, results='asis'}
zv <- Vectorize(z, vectorize.args = c('m', 's', 'n', 'tau'))

@@

hook(zv(.M, .S, .N, .Tau, .C, .MA, .SA, .MB, .SB, .MM), z[.Tau])
mathout(call("=", quote(X["8 df"]^2L), gof))
```

The degrees of freedom result from the 13 predicted mean response times, 
minus 5 free parameters
\(\theta = \langle\mu_{\mathrm A}, \sigma_{\mathrm A}, \mu_{\mathrm B}, \sigma_{\mathrm B}, \mu_{\mathrm M}\rangle\)
(\(c\) is a redundant scaling parameter).

```{r, results="asis", echo=FALSE}
lsq <- function(tau, m, s, n)
{
  invisible(theta <- c(mu_A=0.53, sigma_A=4.3, mu_B=1.34, sigma_B=11.7, mu_M=160))
  optim(theta, gof, tau=tau, m=m, s=s, n=n, hessian=TRUE)
}

@@

```{r params, echo=FALSE}
fit <- lsq(tau, m, s, n)
theta <- fit$par
se <- sqrt(diag(solve(fit$hessian)))
cilo <- qnorm(0.025, theta, se)
ciup <- qnorm(0.975, theta, se)

ntheta <- lapply(lapply(FUN=as.symbol, names(theta)), FUN=mathout, flags=list(cat=FALSE))
ttheta <- lapply(theta, FUN=mathout, flags=list(cat=FALSE))
tcilo <- lapply(cilo, FUN=mathout, flags=list(cat=FALSE))
tciup <- lapply(ciup, FUN=mathout, flags=list(cat=FALSE))

t <- cbind(Parameter=ntheta, Estimate=ttheta, Lo=tcilo, Up=tciup)
knitr::kable(t, caption="Model fit", row.names=FALSE, escape=FALSE)
```

The statistic goodness-of-fit value is \(X^2(8) = 30.54\), \(p = .0002\), 
indicating some lack of fit. Given the large trial numbers in the original
study, some lack of fit is expected. For more detail, especially on fitting the
observed standard deviations, the reader is referred to the original
paper [@schwarz1994].

@@

This package expands the possibilities of R\ by allowing calculations and
graphical rendering of the same terms. Building on current features of R\ and
existing packages for displaying formulas in
R\ [@murrell2000;@allaire2018]), the \pkg{mathml} package bridges the gap
between computational needs, presentation of results, and their reproducibility.
Two possible modes for typesetting are available to export R\ Markdown documents
in different formats or ShinyApp webpages: MathML and MathJax. A further, not-yet-mentioned, use of \pkg{mathml} is related to presentations made using
RMarkdown. By adopting \pkg{mathml} researchers or teachers can use the
RMarkdown to both conduct analysis and show results, smoothing the process and
uniforming the graphical output.

The case study presented above shows that \pkg{mathml} can help to improve
data analyses and statistical reports from an aesthetical perspective, as well
as regarding reproducibility of research. Furthermore, the package may also
allow for a better detection of possible mistakes in R programs.

Similar to most programming languages [@green1977], R\ code is notoriously hard
to read, and the poor legibility of the language is one of the main sources of
mistakes. For illustration, we consider again Equation 10 in
Schwarz [-@schwarz1994].

````{r, results="asis"}
f1 <- function(tau)
{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"]) * 
    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))))
}

@@

The first version has a wrong parenthesis, which is barely visible in the code,
whereas the wrong curly brace in the mathematical representation is immediately
obvious. The correct version is shown below for comparison.

>>>>

---
title: "mathml: Translate R expressions to MathML and LaTeX"
date: "2023-01-21"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the RMarkdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package allows to use the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks allow to extend the package by
  mapping specific variables or functions to new MathML and LaTeX entities. A
  number of examples and a case study illustrate the use of the package in a
  scientific report.
draft: true
author:  
  - name: Matthias Gondan
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: bibliography.bib
preamble:
- \usepackage{cancel}
- \usepackage{amsmath}
Vignette: >
  %\VignetteIndexEntry{mathml: Translate R expressions to MathML and LaTeX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

# Chains: a < b < c

@@

The paper is organized as follows. Section\ 2 presents the technical background
of the package, including the two main classes of Prolog rules for translating
R\ objects to mathematical expressions. Section\ 3 illustrates the main features
of the \pkg{mathml} package, potential issues and workarounds using examples
from the day-to-day perspective of a user. Section\ 4 present a case study for
the use of the package in a scientific report. Section\ 5 concludes with a
discussion and ideas for further development.

@@

The translation of R\ expressions to mathematical output is achieved through a
Prolog interpreter from R\ package \pkg{rolog} [@rolog]. Prolog is a classical logic
programming language with many applications in expert systems, computer
linguistics and symbolic artificial intelligence. The main strength of Prolog is
its concise representation of facts and rules for knowledge and grammar, as well
as its efficient built-in search engine for closed world domains. As it is
well-known, R\ is a statistical programming language for data analysis and
statistical modeling. Whereas Prolog is weak in statistical computation, but
strong in symbolic manipulation, the converse may be said for the R\ language.
The \pkg{rolog} package bridges this gap by providing an interface to a SWI-Prolog
distribution [@swipl] in R. The communication between the two systems is mainly
in the form of queries from R\ to Prolog, but two predicates allow Prolog to
ring back and evaluate terms in R.

@@

As is seen from the output above, the variable `term` is not assigned the result
of the calculation, but an R\ call
[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. Such a call
can eventually be evaluated with `eval()`,

@@

We can include the output in a RMarkdown document by specifying `results='asis'`
in the R\ code chunk, as is done in the next example. The
R\ function `mathout()` has been defined in this vignette; it invokes `mathml()`
for HTML output and `mathjax()` for LaTeX output.

@@

The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
with `N` representing the name of the function and `X` its arguments and
parameters, enclosed in parentheses. A corresponding rule `jax/3` does the same
for MathJax/LaTeX. A list of flags can be used for context-sensitive
translation (see, e.g., the section on errors below).

@@

An expression such as `1 + -2` may be considered unsatisfactory from an
aesthetical perspective. It is correct R\ syntax, though, and is reproduced
accordingly, without the parentheses. Parentheses around negated numbers or
symbols can be added as shown above for `+ (-Inf)`.

Note that an R\ function `hat()` does not exist in base R, it is provided by the
package for convenience and points to the identity function. Quite a few such
convenience functions have been defined
in \pkg{mathml} (see, e.g., next section, and the manual for details).

@@

For multiplications involving only numbers and symbols, the multiplication sign
is omitted. This heuristic does not always produce the desired result;
therefore, \pkg{mathml} defines alternative R\ functions `dot()`, `nodot()`,
and `times()`. These functions calculate a product and produce the respective
multiplication signs. Similarly, `frac()` and `dfrac()` can be used for small
and large fractions.

@@

```{r, results='asis'}
term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
mathout(term)
term <- quote('%+-%'(mean(X), 1.96 * s / sqrt(N))) # functional form of '%+-%'
term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
mathout(term)
```

The example is a reminder that it is not possible to define the precedence of
custom operators in R, and that expressions with such operators are evaluated
strictly from left to right. Again, the problem can be worked around by the
functional form of the operator, or a curly brace to hide the parenthesis but
enforce the correct operator precedence.

@@

There is support for most functions from package _base_, with adequate use and
omission of parentheses.

```{r, results='asis'}
term <- quote(sin(x) + sin(x)^2L + cos(pi/2L) + tan(2L*pi) * expm1(x))
mathout(term)
term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + exp(frac(x, y)))
mathout(term)
```

@@

```{r base-stats, echo=FALSE}
op1 <- list(
  "sin(x)"=quote(sin(x)),
  "cosh(x)"=quote(cosh(x)),
  "tanpi(alpha)"=quote(tanpi(alpha)),
  "asinh(x)"=quote(asinh(x)),
  "log(p)"=quote(log(p)),
  "log1p(x)"=quote(log1p(x)),
  "logb(x,\\ e)"=quote(logb(x, e)),
  "exp(x)"=quote(exp(x)),
  "expm1(x)"=quote(expm1(x)),
  "choose(n,\\ k)"=quote(choose(n, k)),
  "lchoose(n,\\ k)"=quote(lchoose(n, k)),
  "factorial(n)"=quote(factorial(n)), 
  "lfactorial(n)"=quote(lfactorial(n)),
  "sqrt(x)"=quote(sqrt(x)),
  "mean(X)"=quote(mean(X)),
  "abs(x)"=quote(abs(x)))
m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))

@@

For self-written functions, matters are a bit more complicated. For a function
such as `g <- function(...) ...`, the name _g_ is not transparent to R, because
only the function body is represented, as illustrated by the redefinition
of `sign` below.

@@

mathout(sgn)
mathout(call("<-", quote(sgn(x)), sgn))
```

As shown in the example, we can still display functions in the
form `head(x) = body` if we embed the object to be shown into a
call `"<-"(head, body)`.

The function body is generally a nested R\ call of the form `'{'(L)`, with `L`
being a list of commands (the semicolon, not necessary in R, is translated to a
newline). As illustrated in the example, \pkg{mathml} provides limited support
for control structures such as `if`.

@@

For mathematical typesetting in the form
of \(\int f(x)\, dx\), \pkg{mathml} needs to find out the name of the
integration variable. For that purpose, the underlying Prolog bridge provides a
predicate `r_eval/3` that calls R\ from Prolog. In the example above, this
predicate evaluates `formalArgs(args(sin))`, which returns the names of the
arguments of `sin`, namely, `x`.

Note that in the example above, the quoted term is an abbreviation
for `call("integrate", quote(sin), ...)`, with `sin` being an R\ symbol, not a
function. While the R\ function `integrate()` can handle both symbols and
functions, \pkg{mathml} needs the symbol because it is unable to determine the
function name of custom fuctions.

@@

One of R's great features is the possibility to refer to function arguments by
their names, not only by their position in the list of arguments. At the other
end, Prolog does not have such a feature. Therefore, the Prolog handlers for
R\ calls are rather rigid, for example, `integrate/3` accepts exactly three
arguments in a particular order and without names, that
is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print the desired result.

To "canonicalize" function calls with named arguments and arguments in unusual
order, \pkg{mathml} provides an auxiliary R\ function `canonical(f, drop)` that
reorders the argument list of calls to known R\ functions and, 
if `drop=TRUE` (which is the default), also removes the names of the arguments.

@@

Note that the seemingly more convient `term <- quote(A + B)` yields \(A + B\) in
the output---instead of the desired matrix representation. This behavior is
expected because quotation of R calls also quote the components of the
call (here, _A_ and _B_).

## Short mathematical names for R symbols

The hooks can also be used for more complex elements such as R\ calls, with
dotted symbols representing Prolog variables.

```{r, results='asis'}
hook(pbinom(.K, .N, .P), subsupscript(sum(dbinom(i, .N, .P)), i=0L, .K))
mathout(term)
```

@@

We consider the \(t\)-statistic for independent samples with equal variance. To
avoid clutter in the equation, the pooled variance \(s^2_{\mathrm{pool}}\) is
abbreviated, and a comment is given with the expression
for \(s^2_{\mathrm{pool}}\). For this purpose, \pkg{mathml} provides a
function `denote(abbr, expr, info)`, with `expr` actually being
evaluated, `abbr` being rendered, plus a comment of the
form "with `expr` denoting `info`".

@@

The term is evaluated below. `print()` is needed because the return value of an
assignment of the form `t <- dfrac(...)` is not visible in R.

@@

```{r, results='asis'}
t <- quote(dfrac(omit_right(mean(D) - mu[0L]), s / sqrt(N)))
mathout(t, flags=list(error="highlight"))
mathout(t, flags=list(error="fix"))
```

@@

This case study describes a model by Schwarz [-@schwarz1994] from mathematical
psychology using the features of package \pkg{mathml}. Schwarz presents a new
explanation of redundancy gains that occur when observers respond to stimuli of
different sources, and the same information is presented on two or more
channels. In Schwarz' model, decision-making builds on a process of noisy
accumulation of information over time [e.g., @ratcliff2016]. In redundant
stimuli, the model assumes a superposition of channel-specific diffusion
processes that eventually reach an absorbing barrier to elicit the response. For
a detailed description the reader may refer to the original article.

Schwarz' [-@schwarz1994] model refers to two stimuli A and B, presented either
alone or in combination (AB, redundant stimuli), with the redundant stimuli
being presented either simultaneously or with onset asynchrony \(\tau\). The
channel activation is described as a two-dimensional Wiener process with
drifts \(\mu_i\), variances \(\sigma^2_i\), and initial
conditions \(X_i(t = 0) = 0, i = \mathrm{A, B}\). The buildup of 
channel-specific activation may be correlated with \(\rho_{\mathrm{AB}}\), but
we assume \(\rho_{\mathrm{AB}} = 0\) for simplicity.

A response is elicited when the process reaches an absorbing barrier \(c > 0\)
for the first time. In single-target trials, the first passages of \(c\) are
expected at

@@

(One would typically use chunk option `echo=FALSE` to suppress the R code.)
In redundant stimuli, the activations from the channel-specific diffusion
processes add up, \(X_{\mathrm{AB}}(t) = X_{\mathrm A}(t) + X_{\mathrm B}(t)\),
hence the name, superposition. \(X_{\mathrm{AB}}(t)\) is again a Wiener process
with drift \(\mu_{\mathrm A} + \mu_{\mathrm B}\) and
variance \(\sigma^2_{\mathrm A} + \sigma^2_{\mathrm B}\). For the expected
first-passage time, we have

@@

```{r, results="asis"}
ED_async <- function(tau, c, mu_A, sigma_A, mu_B)
{ dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B)) *
    ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
      - (mu_A*tau + c) * exp(dfrac(2L*c*mu_A, sigma_A^2L))
        * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau))))
}

@@

For negative onset asynchrony (i.e., B before A), the parameters are simply
switched.

```{r, results='asis'}
ED <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
{
  if(tau == Inf) return(ED_single(c, mu_A))
  if(tau == -Inf) return(ED_single(c, mu_B))
  if(tau == 0L) return(ED_single(c, mu_A + mu_B))
  if(tau > 0L) return(ED_async(tau, c, mu_A, sigma_A, mu_B))
  if(tau < 0L) return(ED_async(abs(tau), c, mu_B, sigma_B, mu_A))
}

@@

@@

```{r miller-data, echo=FALSE}
tau  <- list(-Inf,-167L,-133L,-100L,-67L,-33L,0L,33L,67L,100L,133L,167L,Inf)
m    <- c(231, 234, 230, 227, 228, 221, 217, 238, 263, 277, 298, 316, 348)
s    <- c( 56,  58,  40,  40,  32,  28,  28,  28,  26,  30,  32,  34,  92)
n    <- c(400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400)

@@

t <- cbind('\\(\\tau\\)'=ttau, '\\(m\\)'=tm, '\\(s\\)'=ts, '\\(n\\)'=tn)
knitr::kable(t, caption="Miller data", row.names=FALSE, escape=FALSE)
```

Assuming that the model is correct, the observable mean reaction times follow an
approximate Normal distribution around the model prediction \(E(T_\tau)\) for
each condition. We can, therefore, use a standard goodness-of-fit measure
by \(z\)-standardisation.

```{r, results='asis'}
z <- function(m, s, n, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
  dfrac(m - denote(mu[tau], ET(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M),
              "the expected mean response time"),
    s / sqrt(n))

@@

The overall goodness-of-fit is the sum of the squared \(z\)-statistics for each
onset asynchrony. Assuming again that the architecture of the model is correct,
but the parameters are adjusted to the data, it follows
a \(\chi^2(8\ \mathrm{df})\)-distribution.

```{r, results='asis'}
zv <- Vectorize(z, vectorize.args = c('m', 's', 'n', 'tau'))
hook(zv(.M, .S, .N, .Tau, .C, .MA, .SA, .MB, .SB, .MM), z[.Tau])

@@

mathout(call("=", quote(X["8 df"]^2L), gof))
```

with the degrees of freedom given by the difference between the number of
observations (`r length(tau)`) and the number of free model
parameters \(\theta = \langle\mu_{\mathrm A}, \sigma_{\mathrm A},
\mu_{\mathrm B}, \sigma_{\mathrm B}, \mu_{\mathrm M}\rangle\); the barrier \(c\)
is only a scaling parameter.

```{r, results="asis", echo=FALSE}
lsq <- function(tau, m, s, n)
{
  invisible(theta <- c(mu_A=0.5, sigma_A=5, mu_B=0.5, sigma_B=5, mu_M=150))
  optim(theta, gof, tau=tau, m=m, s=s, n=n, hessian=TRUE)
}

@@

```{r params, echo=FALSE}
fit <- lsq(tau, m, s, n)
theta <- fit$par
se <- sqrt(diag(solve(fit$hessian)))
ci <- mapply(FUN=qnorm, theta, se, MoreArgs=list(p=c(0.025, 0.975)),
    SIMPLIFY=FALSE)

ntheta <- lapply(lapply(FUN=as.symbol, names(theta)), FUN=mathout,
    flags=list(cat=FALSE))
ttheta <- lapply(theta, FUN=mathout, flags=list(cat=FALSE))
tci <- lapply(ci, FUN=mathout, flags=list(cat=FALSE))

t <- cbind(Parameter=ntheta, Estimate=ttheta, CI=tci)
knitr::kable(t, caption="Model fit", row.names=FALSE, escape=FALSE)
```

The goodness-of-fit statistic indicates some lack of
fit, \(X^2(8\ \mathrm{df}) = `r sprintf("%.2f", fit$value)`, 
  p = `r sprintf("%.4f", 1-pchisq(fit$value, df=8))`\). Given the large trial
numbers in the original study, this is not an unexpected result. For more
detail, especially on fitting the observed standard deviations, the reader is
referred to the original paper [@schwarz1994].

@@

This package allows R\ to render its terms in pretty mathematical equations. It
extends the current features of R\ and existing packages for displaying
mathematical formulas in R\ [@murrell2000;@allaire2018]), bit most
importantly, \pkg{mathml} bridges the gap between computational needs, presentation
of results, and their reproducibility. The package supports both MathML and
Latex/MathJax for use in R\ Markdown documents, presentations and ShinyApp
webpages.

Researchers or teachers can already use RMarkdown to conduct analyses
and show results, and \pkg{mathml} smoothes this process and allows for integrated
calculations and output. As shown in the case study of the previous
section, \pkg{mathml} can help to improve data analyses and statistical reports from
an aesthetical perspective, as well as regarding reproducibility of research.

Furthermore, the package may also allow for a better detection of possible
mistakes in R\ programs. Similar to most programming languages [@green1977],
R\ code is notoriously hard to read, and the poor legibility of the language is
one of the main sources of mistakes. For illustration, we consider again
Equation\ 10 in Schwarz [-@schwarz1994].

````{r, results="asis"}
f1 <- function(tau)
{ dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B) * 
    ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
      - (mu_A*tau + c) * exp(dfrac(2L*mu_A*tau, sigma_A^2L))
        * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau)))))
}

@@

The first version has a wrong parenthesis, which is barely visible in the code,
whereas in the mathematical representation, the wrong curly brace is immediately
obvious (the correct version is shown below for comparison).
